---
title: "GAPMINDER | Child Mortality, GDP per Capita, and Average Daily Income (1960–2026)"
author: ""
format:
  dashboard:
    output-file: index.html
    nav-buttons: [github]
    github: https://github.com/Ta-ver/child_mortality_final_project
    theme: yeti
    embed-resources: true
    css: dashboard-extras.css
    include-in-header:
      - text: |
          <style>
          /* Rounded plot / table cards (bslib chrome). Value boxes: dashboard-extras.css */
          .quarto-dashboard .bslib-card.card:has(.plotly.html-widget),
          .quarto-dashboard .bslib-card.card:has(.reactable.html-widget),
          .quarto-dashboard .bslib-card.card.tabset:has(.tab-content) {
            border-radius: 1.15rem !important;
            overflow: hidden !important;
          }
          .quarto-dashboard .bslib-card.card:has(.plotly.html-widget) .plotly.html-widget,
          .quarto-dashboard .bslib-card.card:has(.reactable.html-widget) .reactable.html-widget {
            border-radius: inherit;
            overflow: hidden;
          }
          </style>
resources:
  - data
  - dashboard-extras.css
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false

# Load packages
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(
  tidyverse,
  here,
  readr,
  plotly,
  reactable,
  htmltools,
  scales,
  sf,
  bslib,
  bsicons,
  rnaturalearth,
  rnaturalearthdata,
  countrycode
)

year_min <- 1960L
year_max <- 2026L
year_cols <- as.character(year_min:year_max)

child_raw <- read_csv(
  here::here("data/child_mortality_0_5_year_olds_dying_per_1000_born.csv"),
  show_col_types = FALSE
)

gdp_raw <- read_csv(here::here("data/gdp_pcap.csv"), show_col_types = FALSE)

inc_raw <- read_csv(here::here("data/mincpcap_cppp.csv"), show_col_types = FALSE)

# Keep 1960–2026 year columns present in each file (intersect avoids missing-column errors)
pick_years <- function(df, id_cols) {
  have <- intersect(year_cols, names(df))
  df |> dplyr::select(all_of(c(id_cols, have)))
}

# Long format: one row per country-year (`pivot_longer`, `mutate`, `rename`)
child <- pick_years(child_raw, "country") |>
  pivot_longer(-country, names_to = "year", values_to = "child_mortality") |>
  mutate(
    year = as.integer(year),
    child_mortality = suppressWarnings(as.numeric(child_mortality))
  )

gdp <- pick_years(gdp_raw, c("geo", "name")) |>
  dplyr::select(-geo) |>
  rename(country = name) |>
  pivot_longer(-country, names_to = "year", values_to = "gdp_per_capita") |>
  mutate(
    year = as.integer(year),
    gdp_per_capita = suppressWarnings(as.numeric(gdp_per_capita))
  )

inc <- pick_years(inc_raw, c("geo", "name")) |>
  dplyr::select(-geo) |>
  rename(country = name) |>
  pivot_longer(-country, names_to = "year", values_to = "avg_daily_income_ppp") |>
  mutate(
    year = as.integer(year),
    avg_daily_income_ppp = suppressWarnings(as.numeric(avg_daily_income_ppp))
  )

# `inner_join`: only country-years present in all three indicators (Gapminder-style alignment)
panel <- child |>
  inner_join(gdp, by = c("country", "year")) |>
  inner_join(inc, by = c("country", "year")) |>
  distinct(country, year, .keep_all = TRUE) |>
  filter(
    if_all(
      c(child_mortality, gdp_per_capita, avg_daily_income_ppp),
      is.finite
    )
  )

global_means <- panel |>
  group_by(year) |>
  summarise(
    child_mortality = mean(child_mortality, na.rm = TRUE),
    gdp_per_capita = mean(gdp_per_capita, na.rm = TRUE),
    avg_daily_income_ppp = mean(avg_daily_income_ppp, na.rm = TRUE),
    .groups = "drop"
  )

snapshot_year <- panel |>
  filter(
    year <= 2026L,
    !is.na(child_mortality),
    !is.na(gdp_per_capita),
    !is.na(avg_daily_income_ppp)
  ) |>
  group_by(year) |>
  summarise(n_countries = n(), .groups = "drop") |>
  filter(n_countries >= 150) |>
  slice_max(year, n = 1) |>
  pull(year)

snap <- panel |>
  filter(year == snapshot_year)

fmt_dollars <- label_dollar(accuracy = 1, big.mark = ",")

# Headline values (child mortality)
mean_child_mort <- mean(snap$child_mortality, na.rm = TRUE)
n_snap_countries <- n_distinct(snap$country)

highest_child_mort_country <- snap |>
  arrange(desc(child_mortality)) |>
  slice_head(n = 1) |>
  pull(country)

highest_child_mort_value <- snap |>
  arrange(desc(child_mortality)) |>
  slice_head(n = 1) |>
  pull(child_mortality) |>
  round(1)

lowest_child_mort_country <- snap |>
  arrange(child_mortality) |>
  slice_head(n = 1) |>
  pull(country)

lowest_child_mort_value <- snap |>
  arrange(child_mortality) |>
  slice_head(n = 1) |>
  pull(child_mortality) |>
  round(1)

# Headline values (GDP per capita)
mean_gdp_cap <- mean(snap$gdp_per_capita, na.rm = TRUE)
highest_gdp_country <- snap |>
  arrange(desc(gdp_per_capita)) |>
  slice_head(n = 1) |>
  pull(country)
highest_gdp_value <- snap |>
  arrange(desc(gdp_per_capita)) |>
  slice_head(n = 1) |>
  pull(gdp_per_capita)
lowest_gdp_country <- snap |>
  arrange(gdp_per_capita) |>
  slice_head(n = 1) |>
  pull(country)
lowest_gdp_value <- snap |>
  arrange(gdp_per_capita) |>
  slice_head(n = 1) |>
  pull(gdp_per_capita)

# Headline values (average daily income)
mean_income_day <- mean(snap$avg_daily_income_ppp, na.rm = TRUE)
highest_income_country <- snap |>
  arrange(desc(avg_daily_income_ppp)) |>
  slice_head(n = 1) |>
  pull(country)
highest_income_value <- snap |>
  arrange(desc(avg_daily_income_ppp)) |>
  slice_head(n = 1) |>
  pull(avg_daily_income_ppp) |>
  round(2)
lowest_income_country <- snap |>
  arrange(avg_daily_income_ppp) |>
  slice_head(n = 1) |>
  pull(country)
lowest_income_value <- snap |>
  arrange(avg_daily_income_ppp) |>
  slice_head(n = 1) |>
  pull(avg_daily_income_ppp) |>
  round(2)
```

```{r}
#| include: false

# World map + top countries chart (child mortality snapshot year)
country_shapes <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

snap_geo <- snap |>
  mutate(
    country_code = countrycode::countrycode(
      country,
      origin = "country.name",
      destination = "iso3c",
      warn = FALSE
    )
  )

child_sf_map <- dplyr::left_join(
  country_shapes,
  snap_geo,
  by = c("adm0_a3" = "country_code")
) |>
  mutate(
    map_label = dplyr::coalesce(country, admin)
  )

child_mort_map_gg <- ggplot(child_sf_map) +
  geom_sf(
    aes(
      fill = child_mortality,
      text = paste0(map_label, ": ", round(child_mortality, 1), " / 1,000")
    )
  ) +
  scale_fill_gradientn(
    colours = c(
      "#0ea5e9",
      "#22d3ee",
      "#a5f3fc",
      "#fde047",
      "#fb923c",
      "#f43f5e"
    ),
    name = "Deaths\nper 1,000",
    na.value = "#e2e8f0"
  ) +
  theme_void() +
  theme(legend.position = "bottom")

child_mort_map_plotly <- ggplotly(child_mort_map_gg, tooltip = "text")

gap_15_child <- snap |>
  arrange(desc(child_mortality)) |>
  slice_head(n = 15) |>
  mutate(
    tooltip_label = paste0(country, ": ", round(child_mortality, 1), " / 1,000")
  )

top_15_child_mort_gg <- ggplot(
  gap_15_child,
  aes(
    y = reorder(country, child_mortality),
    x = child_mortality,
    fill = child_mortality,
    text = tooltip_label
  )
) +
  geom_col() +
  labs(x = "Under-5 deaths per 1,000 live births", y = NULL) +
  theme_minimal() +
  theme(legend.position = "none")

top_15_child_mort_ggplotly <- ggplotly(top_15_child_mort_gg, tooltip = "text")
```

# **Child Mortality**

## Row 1 {height="30%"}

```{r}
value_box(
  title = "Highest child mortality",
  value = paste0(highest_child_mort_value, " /1k"),
  showcase = bsicons::bs_icon("graph-up-arrow"),
  theme = value_box_theme(bg = "#e74c3c"),
  p(paste0("(", highest_child_mort_country, ")"))
)
```

```{r}
value_box(
  title = "Lowest child mortality",
  value = paste0(lowest_child_mort_value, " /1k"),
  showcase = bsicons::bs_icon("graph-down-arrow"),
  theme = value_box_theme(bg = "#27ae60"),
  p(paste0("(", lowest_child_mort_country, ")"))
)
```

```{r}
value_box(
  title = paste0("Mean child mortality (", snapshot_year, ")"),
  value = paste0(round(mean_child_mort, 1), " /1k"),
  showcase = bsicons::bs_icon("heart-pulse"),
  theme = value_box_theme(bg = "#c0392b"),
  p(paste0("(", n_snap_countries, " countries; deaths per 1,000 born)"))
)
```

## Row 2 {height="70%"}

###  {width="65%"}

```{r}
cat("title=", paste0("World map: under-5 mortality (", snapshot_year, ")"))
child_mort_map_plotly
```

###  {.tabset}

#### Top 15 Countries CMR

```{r}
cat("title=", paste0("Top 15 countries (", snapshot_year, ")"))
top_15_child_mort_ggplotly
```

#### Global Average Trend

```{r}
#| title: "Global average trend (simple mean across countries)"

p_cm_global <- global_means |>
  plot_ly(
    x = ~year,
    y = ~child_mortality,
    type = "scatter",
    mode = "lines",
    line = list(color = "#c0392b", width = 3),
    hovertemplate = "Year: %{x}<br>Mean: %{y:.1f} deaths / 1,000<extra></extra>"
  ) |>
  layout(
    title = list(text = "Mean child mortality over time (unweighted country average)", font = list(size = 13)),
    xaxis = list(title = "Year"),
    yaxis = list(title = "Mean under-5 deaths per 1,000 live births"),
    margin = list(l = 60, r = 20, t = 48, b = 50)
  )

p_cm_global
```

#### Distribution

```{r}
plot_ly(
  snap,
  x = ~child_mortality,
  type = "histogram",
  nbinsx = 35,
  marker = list(color = "#c0392b", line = list(color = "white", width = 0.5)),
  hovertemplate = "Bin center: %{x}<br>Number of countries: %{y}<extra></extra>"
) |>
  layout(
    title = list(text = paste0("Distribution of child mortality (", snapshot_year, ")"), font = list(size = 13)),
    xaxis = list(title = "Under-5 deaths per 1,000 live births"),
    yaxis = list(title = "Number of countries"),
    margin = list(l = 60, r = 20, t = 48, b = 50)
  )
```

# **GDP per Capita**

## Row 1 {height="30%"}

###  {width="38%"}

```{r}
value_box(
  title = "Highest GDP per capita",
  value = fmt_dollars(highest_gdp_value),
  showcase = bsicons::bs_icon("graph-up-arrow"),
  theme = value_box_theme(bg = "#2980b9"),
  p(paste0("(", highest_gdp_country, ")"))
)
```

###  {width="31%"}

```{r}
value_box(
  title = "Lowest GDP per capita",
  value = fmt_dollars(lowest_gdp_value),
  showcase = bsicons::bs_icon("graph-down-arrow"),
  theme = value_box_theme(bg = "#7f8c8d"),
  p(paste0("(", lowest_gdp_country, ")"))
)
```

###  {width="31%"}

```{r}
value_box(
  title = paste0("Mean GDP per capita (", snapshot_year, ")"),
  value = fmt_dollars(mean_gdp_cap),
  showcase = bsicons::bs_icon("currency-dollar"),
  theme = value_box_theme(bg = "#3498db"),
  p("PPP international dollars")
)
```

## Row 2 {height="70%"}

###  {width="60%"}

```{r}
#| title: "Global average GDP per capita (PPP)"

p_gdp_global <- global_means |>
  plot_ly(x = ~year, y = ~gdp_per_capita, type = "scatter", mode = "lines",
          line = list(color = "#2980b9", width = 3),
          hovertemplate = "Year: %{x}<br>Mean GDP/cap: $%{y:,.0f}<extra></extra>") |>
  layout(
    title = list(text = "Mean GDP per capita over time (unweighted country average)", font = list(size = 13)),
    xaxis = list(title = "Year"),
    yaxis = list(title = "Mean GDP per capita (PPP international $)"),
    margin = list(l = 70, r = 20, t = 48, b = 50)
  )

p_gdp_global
```

###  {.tabset}

#### Top 15 economies

```{r}
top15_g <- snap |>
  slice_max(gdp_per_capita, n = 15, with_ties = FALSE) |>
  mutate(country = fct_reorder(country, gdp_per_capita))

plot_ly(
  top15_g,
  x = ~gdp_per_capita,
  y = ~country,
  type = "bar",
  orientation = "h",
  marker = list(color = "#2980b9"),
  hovertemplate = "%{y}<br>GDP per capita: $%{x:,.0f}<extra></extra>"
) |>
  layout(
    title = list(text = paste0("Top 15 economies by GDP per capita (", snapshot_year, ")"), font = list(size = 13)),
    xaxis = list(title = "GDP per capita (PPP international $)"),
    yaxis = list(title = "", categoryorder = "total ascending"),
    margin = list(l = 140, r = 20, t = 48, b = 50)
  )
```

#### Distribution

```{r}
plot_ly(
  snap,
  x = ~gdp_per_capita,
  type = "histogram",
  nbinsx = 40,
  marker = list(color = "#2980b9", line = list(color = "white", width = 0.5)),
  hovertemplate = "Bin center: $%{x}<br>Number of countries: %{y}<extra></extra>"
) |>
  layout(
    title = list(text = paste0("Distribution of GDP per capita (", snapshot_year, ")"), font = list(size = 13)),
    xaxis = list(title = "GDP per capita (PPP $)"),
    yaxis = list(title = "Number of countries"),
    margin = list(l = 60, r = 20, t = 48, b = 50)
  )
```

# **Average Daily Income**

## Row 1 {height="35%"}

```{r}
value_box(
  title = "Highest average daily income",
  value = paste0("$", highest_income_value),
  showcase = bsicons::bs_icon("graph-up-arrow"),
  theme = value_box_theme(bg = "#16a085"),
  p(paste0("(", highest_income_country, ")"))
)
```

```{r}
value_box(
  title = "Lowest average daily income",
  value = paste0("$", lowest_income_value),
  showcase = bsicons::bs_icon("graph-down-arrow"),
  theme = value_box_theme(bg = "#f39c12"),
  p(paste0("(", lowest_income_country, ")"))
)
```

```{r}
value_box(
  title = paste0("Mean income per day (", snapshot_year, ")"),
  value = paste0("$", round(mean_income_day, 1)),
  showcase = bsicons::bs_icon("cash-stack"),
  theme = value_box_theme(bg = "#1abc9c"),
  p("PPP dollars per day")
)
```

## Row 2

###  {width="60%"}

```{r}
#| title: "Global average daily income (PPP $ / day)"

p_inc_global <- global_means |>
  plot_ly(
    x = ~year,
    y = ~avg_daily_income_ppp,
    type = "scatter",
    mode = "lines",
    line = list(color = "#16a085", width = 3),
    hovertemplate = "Year: %{x}<br>Mean income: $%{y:.2f} / day<extra></extra>"
  ) |>
  layout(
    title = list(text = "Mean daily income over time (unweighted country average)", font = list(size = 13)),
    xaxis = list(title = "Year"),
    yaxis = list(title = "Mean PPP dollars per day"),
    margin = list(l = 60, r = 20, t = 48, b = 50)
  )

p_inc_global
```

###  {.tabset}

#### Top 10 vs Lowest 10 countries

```{r}
# Top 10 and lowest 10 by PPP $/day (snapshot year), IMR-style dual strip
income_low10 <- snap |>
  slice_min(avg_daily_income_ppp, n = 10, with_ties = FALSE) |>
  arrange(avg_daily_income_ppp) |>
  mutate(
    band = "Lowest 10",
    y_label = paste0(country, " (", round(avg_daily_income_ppp, 2), " $/day)")
  )

income_top10 <- snap |>
  slice_max(avg_daily_income_ppp, n = 10, with_ties = FALSE) |>
  arrange(avg_daily_income_ppp) |>
  mutate(
    band = "Top 10",
    y_label = paste0(country, " (", round(avg_daily_income_ppp, 2), " $/day)")
  )

income_compare <- dplyr::bind_rows(income_low10, income_top10) |>
  mutate(
    y_label = forcats::fct_inorder(y_label),
    hover_txt = paste0(
      country, "<br>",
      band, "<br>",
      "PPP income: $", round(avg_daily_income_ppp, 2), " / day"
    )
  )

y_levels <- levels(income_compare$y_label)

plot_ly(
  income_compare,
  x = ~avg_daily_income_ppp,
  y = ~y_label,
  color = ~band,
  colors = c("Top 10" = "#0d9488", "Lowest 10" = "#ea580c"),
  type = "bar",
  orientation = "h",
  text = ~hover_txt,
  hoverinfo = "text"
) |>
  layout(
    title = list(
      text = paste0("Top 10 vs lowest 10 by daily income (", snapshot_year, ")"),
      font = list(size = 13)
    ),
    xaxis = list(title = "Average daily income (PPP $ per day)"),
    yaxis = list(
      title = "",
      categoryorder = "array",
      categoryarray = y_levels
    ),
    legend = list(title = "", orientation = "h", x = 0, y = 1.08),
    margin = list(l = 24, r = 24, t = 56, b = 48)
  ) |>
  config(displayModeBar = TRUE, displaylogo = FALSE)
```

#### Distribution

```{r}
plot_ly(
  snap,
  x = ~avg_daily_income_ppp,
  type = "histogram",
  nbinsx = 40,
  marker = list(color = "#16a085", line = list(color = "white", width = 0.5)),
  hovertemplate = "Bin center: $%{x}<br>Number of countries: %{y}<extra></extra>"
) |>
  layout(
    title = list(text = paste0("Distribution of daily income (", snapshot_year, ")"), font = list(size = 13)),
    xaxis = list(title = "PPP dollars per day"),
    yaxis = list(title = "Number of countries"),
    margin = list(l = 60, r = 20, t = 48, b = 50)
  )
```

# **Cross-indicator Views**

## Row 1 {height="420px"}

### {width=50%}

::: {.card title="Key Takeaways"}
-   **Wealth vs. survival:** Countries with higher log GDP per capita or higher daily income usually show **lower** under-5 mortality, but the cloud is wide—policy, inequality, and conflict matter beyond averages.
-   **GDP deciles:** Mean child mortality **falls** from the lowest to the highest GDP decile at the snapshot year, while mean income rises with the same grouping - purely descriptive aggregation.
-   **Global lines:** The three series scaled to their own maxima show **parallel long-run movement**: GDP and income rise while mean child mortality falls (unweighted country means each year).
:::

### {width=50%}

```{r}
cat("title=", paste0("Interactive Country Table (", snapshot_year, ")"))

reactable::reactable(
  snap |>
    dplyr::select(country, child_mortality, gdp_per_capita, avg_daily_income_ppp) |>
    dplyr::arrange(dplyr::desc(gdp_per_capita)),
  searchable = TRUE,
  striped = TRUE,
  highlight = TRUE,
  defaultPageSize = 12,
  columns = list(
    country = reactable::colDef(name = "Country", minWidth = 140),
    child_mortality = reactable::colDef(
      name = "Child mortality (/1k)",
      format = reactable::colFormat(digits = 1)
    ),
    gdp_per_capita = reactable::colDef(
      name = "GDP per capita (PPP $)",
      format = reactable::colFormat(digits = 0)
    ),
    avg_daily_income_ppp = reactable::colDef(
      name = "Daily income ($/day)",
      format = reactable::colFormat(digits = 2)
    )
  )
)
```

## Row {height="480px"}

```{r}
#| title: "Wealth vs Child survival"

continent_lvls <- c("Africa", "Americas", "Asia", "Europe", "Oceania", "Unknown")
continent_cols <- c(
  "Africa" = "#c0392b",
  "Americas" = "#2980b9",
  "Asia" = "#d68910",
  "Europe" = "#8e44ad",
  "Oceania" = "#16a085",
  "Unknown" = "#7f8c8d"
)

wealth_frames <- panel |>
  filter(year >= year_min, year <= year_max) |>
  mutate(
    continent = countrycode::countrycode(
      country,
      origin = "country.name",
      destination = "continent",
      warn = FALSE
    ),
    continent = factor(
      dplyr::coalesce(continent, "Unknown"),
      levels = continent_lvls
    ),
    log_gdp = log10(pmax(gdp_per_capita, 1)),
    text = paste0(
      country, " (", year, ")<br>",
      "Continent: ", as.character(continent), "<br>",
      "Mortality: ", round(child_mortality, 1), " / 1,000<br>",
      "GDP/cap: ", fmt_dollars(gdp_per_capita), "<br>",
      "Income: $", round(avg_daily_income_ppp, 2), " / day"
    )
  )

x_rng <- range(wealth_frames$log_gdp, na.rm = TRUE)
y_rng <- range(wealth_frames$child_mortality, na.rm = TRUE)
pad_x <- max(diff(x_rng) * 0.03, 0.02)
pad_y <- max(diff(y_rng) * 0.03, 1)
max_inc <- max(wealth_frames$avg_daily_income_ppp, na.rm = TRUE)
sizeref_val <- max(2 * max_inc / 40^2, 0.08)

plot_ly(
  wealth_frames,
  x = ~log_gdp,
  y = ~child_mortality,
  size = ~avg_daily_income_ppp,
  color = ~continent,
  colors = continent_cols,
  frame = ~year,
  text = ~text,
  type = "scatter",
  mode = "markers",
  marker = list(
    sizemode = "diameter",
    sizeref = sizeref_val,
    line = list(width = 0.4, color = "white")
  ),
  hoverinfo = "text"
) |>
  layout(
    title = list(
      text = paste0(
        "log10(GDP per capita) vs under-5 mortality (",
        year_min, "–", year_max, ", animated by year)"
      ),
      font = list(size = 14)
    ),
    xaxis = list(
      title = "log10(GDP per capita, PPP $)",
      range = c(x_rng[1] - pad_x, x_rng[2] + pad_x)
    ),
    yaxis = list(
      title = "Under-5 deaths per 1,000 live births",
      range = c(max(0, y_rng[1] - pad_y), y_rng[2] + pad_y)
    ),
    margin = list(l = 70, r = 20, t = 56, b = 50),
    showlegend = TRUE,
    legend = list(title = list(text = "Continent"), orientation = "h", y = 1.12)
  ) |>
  animation_opts(frame = 120, transition = 0, redraw = TRUE) |>
  animation_slider(
    currentvalue = list(prefix = "Year ", font = list(size = 14))
  ) |>
  config(displayModeBar = TRUE, displaylogo = FALSE)
```

## Row {height="480px"}

### Daily income vs Mortality

```{r}
b2 <- snap |>
  mutate(
    text = paste0(
      country, "<br>",
      "Income: $", round(avg_daily_income_ppp, 2), " / day<br>",
      "Mortality: ", round(child_mortality, 1), " / 1,000"
    )
  )

plot_ly(
  b2,
  x = ~avg_daily_income_ppp,
  y = ~child_mortality,
  color = ~gdp_per_capita,
  text = ~text,
  type = "scatter",
  mode = "markers",
  marker = list(size = 10, line = list(width = 0.3, color = "white")),
  hoverinfo = "text"
) |>
  layout(
    title = list(
      text = paste0("Daily income vs child mortality (", snapshot_year, ")"),
      font = list(size = 14)
    ),
    xaxis = list(title = "Average daily income (PPP $ per day)"),
    yaxis = list(title = "Under-5 deaths per 1,000 live births"),
    margin = list(l = 70, r = 20, t = 56, b = 50)
  ) |>
  colorbar(title = "GDP/cap<br>PPP $")
```

### Child mortality by GDP decile

```{r}
dec <- snap |>
  filter(!is.na(gdp_per_capita)) |>
  mutate(decile = ntile(gdp_per_capita, 10)) |>
  group_by(decile) |>
  summarise(
    child_mortality = mean(child_mortality, na.rm = TRUE),
    avg_daily_income_ppp = mean(avg_daily_income_ppp, na.rm = TRUE),
    gdp_per_capita = mean(gdp_per_capita, na.rm = TRUE),
    .groups = "drop"
  )

plot_ly(
  dec,
  x = ~decile,
  y = ~child_mortality,
  type = "bar",
  marker = list(color = "#c0392b"),
  hovertemplate = "Decile %{x}<br>Mean deaths / 1,000: %{y:.1f}<extra></extra>"
) |>
  layout(
    title = list(
      text = paste0("Mean child mortality by GDP decile (", snapshot_year, ")"),
      font = list(size = 13)
    ),
    xaxis = list(title = "GDP per capita decile (1 = lowest, 10 = highest)", dtick = 1),
    yaxis = list(title = "Mean under-5 deaths per 1,000 live births", rangemode = "tozero"),
    margin = list(l = 70, r = 20, t = 48, b = 50)
  )
```

### Mean GDP per capita by decile

```{r}
plot_ly(
  dec,
  x = ~decile,
  y = ~gdp_per_capita,
  type = "bar",
  marker = list(color = "#2980b9"),
  hovertemplate = "Decile %{x}<br>Mean GDP/cap: $%{y:,.0f}<extra></extra>"
) |>
  layout(
    title = list(
      text = paste0("Mean GDP per capita by decile (", snapshot_year, ")"),
      font = list(size = 13)
    ),
    xaxis = list(title = "GDP per capita decile (1 = lowest)", dtick = 1),
    yaxis = list(title = "Mean GDP per capita (PPP $)", rangemode = "tozero"),
    margin = list(l = 70, r = 20, t = 48, b = 50)
  )
```

### Mean daily income by decile

```{r}
plot_ly(
  dec,
  x = ~decile,
  y = ~avg_daily_income_ppp,
  type = "bar",
  marker = list(color = "#16a085"),
  hovertemplate = "Decile %{x}<br>Mean income: $%{y:.2f} / day<extra></extra>"
) |>
  layout(
    title = list(
      text = paste0("Mean daily income by GDP decile (", snapshot_year, ")"),
      font = list(size = 13)
    ),
    xaxis = list(title = "GDP per capita decile (1 = lowest)", dtick = 1),
    yaxis = list(title = "Mean PPP dollars per day", rangemode = "tozero"),
    margin = list(l = 70, r = 20, t = 48, b = 50)
  )
```

## Row {height="520px"}

```{r}
#| title: "How three global averages moved together"

p3 <- global_means |>
  pivot_longer(
    c(child_mortality, gdp_per_capita, avg_daily_income_ppp),
    names_to = "indicator",
    values_to = "value"
  ) |>
  mutate(
    indicator = recode(
      indicator,
      child_mortality = "Child mortality (/1,000)",
      gdp_per_capita = "GDP per capita (PPP $)",
      avg_daily_income_ppp = "Avg daily income ($/day)"
    )
  ) |>
  group_by(indicator) |>
  mutate(
    value_scaled = value / max(value, na.rm = TRUE),
    htext = paste0(indicator, "<br>Year: ", year, "<br>Scaled to series max: ", round(value_scaled, 3))
  ) |>
  ungroup()

plot_ly(
  p3,
  x = ~year,
  y = ~value_scaled,
  color = ~indicator,
  text = ~htext,
  type = "scatter",
  mode = "lines",
  hovertemplate = "%{text}<extra></extra>"
) |>
  layout(
    title = list(
      text = "Global averages: each series scaled to its own maximum (= 1)",
      font = list(size = 14)
    ),
    xaxis = list(title = "Year"),
    yaxis = list(title = "Scaled level (1 = series peak)", rangemode = "tozero"),
    legend = list(orientation = "h", x = 0, y = -0.25),
    margin = list(l = 70, r = 20, t = 56, b = 80)
  )
```

# Download

Below is the link to download the CSV files used in this dashboard.

```{r}
#| title: "Dataset downloads"

tags$div(
  class = "card card-body bg-light",
  tags$h5(class = "card-title", "Dataset downloads"),
  tags$ul(
    tags$li(HTML(paste0(
      '<a href="data/child_mortality_0_5_year_olds_dying_per_1000_born.csv" download>child_mortality_0_5_year_olds_dying_per_1000_born.csv</a> - ',
      "under-5 deaths per 1,000 live births by country and year."
    ))),
    tags$li(HTML(paste0(
      '<a href="data/gdp_pcap.csv" download>gdp_pcap.csv</a> - ',
      "GDP per capita in international dollars (PPP), inflation-adjusted."
    ))),
    tags$li(HTML(paste0(
      '<a href="data/mincpcap_cppp.csv" download>mincpcap_cppp.csv</a> - ',
      "average daily income in PPP dollars per day."
    )))
  ),
  tags$p(
    class = "text-muted small mb-0")
)
```

# About

This dashboard, created by Tavershima Adongo, uses harmonised country-year observations from the three indicators below:

**Child mortality (0–5)** counts how many children die before their fifth birthday per 1,000 live births - lower values mean better child survival.

**GDP per capita** summarises average economic output per person in a year, expressed in purchasing-power-parity (PPP) international dollars so comparisons across countries are less distorted by market exchange rates.

**Average daily income** (also PPP) describes typical material living standards in dollars per day; it moves with productivity and distribution, not only with GDP.

Comparing these indicators matters because wealth and income are strongly associated with nutrition, sanitation, health systems, and maternal education - all of which shape child survival. The relationship is not perfectly deterministic: policy choices, inequality, conflict, and measurement differences create scatter around any trend. Still, viewing **child mortality**, **GDP per capita**, and **average daily income** together highlights how economic resources map (imperfectly) onto human outcomes, and how global progress over 1960 - 2026 has unfolded in parallel across these dimensions.

**Note on averages:** line charts labelled “global average” use a **simple mean across countries** in each year (each country weighted equally), not a population-weighted world average.

**Cross-indicator page:** deciles split countries at a single snapshot year (`r snapshot_year`) by GDP per capita. A small number of economies appear in the mortality file but not in the GDP/income tables; those rows are excluded from merged views.

